---
title: "Introduction to R and Quarto"
toc-expand: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

In this course you will use **R** to analyse income and tax data, and
**Quarto** to write up your results. This page introduces both. It is
meant to be read at a keyboard: open RStudio and type the commands in as
you go. Reading code is not the same as writing it.

No prior programming experience is assumed.

::: callout-important
## Before you start: download two files

Download both of these now, and save them in the **same folder** — for
example a folder called EC313 on your Desktop.

[{{< fa solid file-export >}} `ec313cis.csv`](/files/tutorials/ec313cis.csv){download="ec313cis.csv"}
— the data used throughout this page

[{{< fa solid file-export >}} `ec313-r-template.qmd`](/files/tutorials/ec313-r-template.qmd){download="ec313-r-template.qmd"}
— a practice Quarto document that runs the main steps on this page
:::

## The three pieces

It helps to be clear about what each piece of software does, because
students often confuse them.

| Software | What it is |
|:-----------------|:-----------------------------------------------------|
| R | The statistical language. It does the actual computing. |
| RStudio | The program you work in. It is a comfortable front end for R. |
| Quarto | The document system. It turns your code and writing into a PDF. |

An analogy: R is the engine, RStudio is the dashboard and steering
wheel, and Quarto is the report you hand in describing where you drove.

Install R from <https://cran.r-project.org> and RStudio from
<https://posit.co/download/rstudio-desktop/>. Install R first. Quarto
comes bundled with RStudio, so you do not need to download it
separately.

# Part 1: R and RStudio

This part is about R itself — the language you write, and the program
you write it in. Work through it at a keyboard. By the end you will be
able to load a dataset, describe it, reshape it, graph it, and run a
regression.

## Finding your way around RStudio

You installed two programs, but from now on you only ever open one of
them. **Opening RStudio starts R for you**, running quietly in the
background. You never need to open R itself, and most people never do.

You can watch this happen. The moment RStudio finishes loading, the
Console prints R's version number and startup message — that is R
announcing itself, already running and waiting for instructions. To go
back to the analogy: you get into the car, and the engine is already
turning over.

When you open RStudio you will see four panes. They can be rearranged
and re-coloured, so yours may not look identical to the picture, but
every installation has these same four.

![RStudio just after opening a project, before you have done anything.
The panes fill up as you work.](/files/img/rstudio-interface.png){#fig-rstudio
fig-alt="The RStudio window divided into four panes: an empty script at
top left, the Console showing the R startup message at bottom left, an
empty Environment pane at top right, and the Files pane at bottom
right."}

1.  **Source** (top left) — where you write and save your code. In the
    picture it holds an empty script called `Untitled1`. When you open a
    Quarto document here, a **Render** button appears at the top of this
    pane.
2.  **Console** (bottom left) — where R actually runs. Code you send
    from Source appears here along with its results, and you can type
    directly at the `>` prompt for quick experiments.
3.  **Environment** (top right) — a list of every object you have
    created, with its size. It says "Environment is empty" in the
    picture because nothing has been run yet. Once you load data, this
    is where you check that it really has the 500 rows you expected.
4.  **Files / Plots / Packages / Help** (bottom right) — in the picture
    the **Files** tab is showing the contents of the project folder.
    Your graphs appear under **Plots**, and help pages under **Help**.

::: callout-important
## The one habit that matters most

Write your code in the **Source** pane and save it. Code typed into the
Console disappears. Every assignment in this course requires you to
submit your code, so if it is not saved in a file, you do not have it.
:::

## R basics

Everything you work with in R is an **object**: a named thing that holds
something. A single income is an object. A column of 50,000 incomes is
an object. A whole survey, a graph, the results of a regression — all
objects. You give each one a name, and it appears in the Environment
pane the moment you create it.

What distinguishes one kind of object from another is what it holds and
how much of it. The rest of this section introduces them in order of
size, starting with the smallest.

### Scalars and vectors

R works by storing things in named **objects**. You create one with the
assignment arrow `<-`:

```{r}
income <- 52000
income
```

The name goes on the left, the value on the right. An object holding a
single value like this is called a **scalar**. Once it exists, you can
compute with it:

```{r}
taxrate <- 0.15
income * taxrate
```

Names are case sensitive: `income` and `Income` are different objects.
This is the single most common source of beginner errors.

Most of the time you want more than one value. A **vector** holds
several values of the same type, in order, and you build one with `c()`,
which is short for "combine":

```{r}
incomes <- c(52000, 61000, 44500, 89000, 120000)
incomes
```

Vectors are the fundamental data structure in R. A variable in a
dataset — everyone's total income, everyone's age — is a vector.

#### Vectors have a type

Every element of a vector must be the same type. The three you will meet
constantly:

```{r}
taxpaid <- c(8500.50, 12400.00, 3150.25)   # numeric
person  <- c("Ali", "Bea", "Cal")          # character (text, in quotes)
filed   <- c(TRUE, FALSE, TRUE)            # logical
```

`class()` tells you which you have:

```{r}
class(taxpaid)
class(person)
class(filed)
```

If you mix types, R quietly converts everything to the most
accommodating one rather than complaining:

```{r}
c(1, "two", 3)
```

Everything became text — note the quotation marks. This is worth
remembering, because a numeric variable that arrives as text will refuse
to average, and this is usually why. When it happens, convert it
deliberately with `as.numeric()`:

```{r}
as.numeric(c("8500.50", "12400.00"))
```

`as.numeric()` also turns a logical vector into numbers, with `TRUE`
becoming 1 and `FALSE` becoming 0. This is the usual way to make a
**dummy variable**:

```{r}
filed
as.numeric(filed)
```

### Data frames

#### A dataset is vectors side by side

A single vector holds one variable. Real data has many variables
measured on the same people, and we need to keep them lined up: the
third age must belong to the same person as the third income.

That is all a **data frame** is — a set of vectors of equal length,
standing side by side as columns. Each column is one variable. Each row
is one observation, usually one person.

We can build a small one by hand to see the structure. `data.frame()`
takes named vectors and stacks them into columns:

```{r}
people <- data.frame(
  name   = c("Ali", "Bea", "Cal", "Dee", "Eli"),
  age    = c(24, 38, 51, 29, 45),
  educ   = c("High school", "University", "College", "University", "High school"),
  totinc = c(31000, 74000, 52500, 66000, 43000)
)

people
```

Five people, four variables. Notice that the columns have different
types — `name` and `educ` are text, `age` and `totinc` are numeric — but
*within* a column the type is the same, exactly as with a vector.

Every dataset you meet in this course has this shape. The data used
later on this page, and the data in your assignments, are the same thing
with more rows and more columns.

#### Columns are vectors

The `$` operator pulls one column out of a data frame, and what comes
back is an ordinary vector:

```{r}
people$totinc
```

This is the connection worth holding on to. A data frame is not a new
kind of thing to learn — it is a tidy arrangement of the vectors you
already understand.

#### Picking out elements

Square brackets select elements of a vector by position:

```{r}
incomes[1]
incomes[2:3]
```

More useful, you can select by condition:

```{r}
incomes > 60000
incomes[incomes > 60000]
```

The first line returns a logical vector answering the question for each
element; the second keeps only the elements where the answer was `TRUE`.

The other comparisons work the same way: `>=` means greater than or
equal to, `<` and `<=` are less than (or equal to), `==` is equal to,
and `!=` is not equal to:

```{r}
person != "Bea"
```

Because a column of a data frame is just a vector, the very same
brackets work on it:

```{r}
people$totinc[2]
people$totinc[people$totinc > 50000]
```

Picking out whole rows and columns of a data frame works a little
differently, and we will do it with two much friendlier functions,
`filter()` and `select()`, once the tidyverse is loaded.

#### Dates

Dates have their own type. `as.Date()` turns text written as
year-month-day into a date:

```{r}
deadline <- as.Date("2025-04-30")
deadline
class(deadline)
```

Dates can be compared just like numbers, with later dates counting as
larger. Which of these three tax returns were filed by the deadline?

```{r}
filed_on <- as.Date(c("2025-03-15", "2025-04-30", "2025-05-12"))
filed_on <= deadline
```

When you read a data file, a column written as dates, such as
2025-04-30, usually arrives as a date already, so you can compare it
with `as.Date()` straight away.

### Other kinds of object

Vectors and data frames are the two you will use constantly, and between
them they cover almost everything in this course.

R does have other kinds of object, and a few of them will turn up later
once we start running regressions. There is no need to learn them in
advance — we will deal with each one when it appears and there is
something useful to do with it.

### Functions

Objects hold your data. **Functions** are what you do to them. You call
one by writing its name followed by parentheses, with the thing you want
it to work on inside:

```{r}
sqrt(16)
round(24.567, 1)
```

You have already used one: `c()` is a function that takes a set of
values and returns a vector.

The things inside the parentheses are called **arguments**. Above,
`round()` took two: the number to round, and how many decimal places to
keep. Many arguments are optional and have sensible defaults — writing
`round(24.567)` alone rounds to a whole number.

Functions that summarise data take a whole vector and return a single
number:

```{r}
mean(incomes)
median(incomes)
length(incomes)
```

::: callout-tip
## Getting help

Typing `?round` in the Console opens the help page for a function, which
lists every argument it accepts. Help pages are dense, but scroll to the
bottom: there are almost always worked examples.
:::

#### Missing values

Arguments matter more than they first appear. Real survey data is full
of gaps, which R writes as `NA`, meaning "not available." R refuses to
guess what a missing value should be:

```{r}
incomes_missing <- c(52000, 61000, NA, 89000, 120000)
mean(incomes_missing)
```

The answer is `NA`, because the true mean genuinely is unknown. To
average the values you do have, say so explicitly with an argument:

```{r}
mean(incomes_missing, na.rm = TRUE)
```

::: callout-warning
Getting `NA` where you expected a number almost always means missing
values in your data, not a mistake in your code. Check with
`sum(is.na(x))` to see how many there are before deciding what to do.
:::

### Packages

R comes with a great deal built in, but most of the convenient tools
live in **packages**: bundles of extra functions that somebody has
written and shared. There are thousands of them, and they are free.

Installing a package downloads it onto your computer. You do this once,
in the Console:

```{r}
#| eval: false
install.packages("packagename")
```

Loading a package makes its functions available for use. You do this
every time you start a new session, at the top of your document:

```{r}
#| eval: false
library(packagename)
```

::: callout-note
`install.packages()` is like buying a book; `library()` is like taking
it off the shelf. You buy once, but you take it off the shelf every
time you sit down to work.
:::

### Base R and the tidyverse

Everything we have done so far — `c()`, `class()`, `data.frame()`,
square brackets — is **base R**: the functions that come with R itself,
no packages needed.

Base R can do everything this course requires, but a lot of it is
awkward. The **tidyverse** is a set of packages whose functions are
built specifically for working with data, and they make the common jobs
far easier to write and to read. Three of them do most of the work:

| Package | What its functions are for |
|:-------------|:----------------------------------------------------|
| **readr** | Getting data from a file into R |
| **dplyr** | Working with data once it is there: keeping certain rows, creating new variables, computing averages by group |
| **ggplot2** | Turning data into graphs |

You will meet all three in the sections that follow. They are built to a
common design, so the habits you pick up with one carry over to the
others, and they are meant to be chained together — the output of one
function becomes the input of the next.

The convenient part is that you do not install them one at a time.
Installing the tidyverse installs the whole set, and a single
`library()` call loads them all.

Install it once:

```{r}
#| eval: false
install.packages("tidyverse")
```

And load it now, since everything below depends on it:

```{r}
#| message: false
library(tidyverse)
```

::: callout-tip
When you search for help online you will find answers written in both
styles, sometimes for the same question. Neither is wrong. Use whichever
you understand, but expect the code in lectures and assignments to look
like the tidyverse.
:::

## Working with a realistic dataset

```{r}
#| label: make-cis
#| include: false
set.seed(313)

n <- 500

cis <- tibble(
  id     = 1:n,
  age    = sample(20:64, n, replace = TRUE),
  female = rbinom(n, 1, 0.5),
  educ   = sample(
    c("Less than HS", "High school", "College", "University"),
    n, replace = TRUE, prob = c(0.10, 0.30, 0.30, 0.30)
  )
) |>
  mutate(
    yrschool = case_when(
      educ == "Less than HS" ~ 10,
      educ == "High school"  ~ 12,
      educ == "College"      ~ 14,
      educ == "University"   ~ 16
    ),
    totinc = round(
      exp(8.80 + 0.09 * yrschool + 0.048 * age -
            0.00050 * age^2 - 0.15 * female + rnorm(n, 0, 0.55))
    ),
    transfers = round(pmax(0, 9000 - 0.06 * totinc + rnorm(n, 0, 1200))),
    inctax = round(
      (0.20 * pmax(totinc - 20000, 0) +
         0.10 * pmax(totinc - 55000, 0) +
         0.10 * pmax(totinc - 110000, 0) +
         0.05 * pmax(totinc - 170000, 0)) *
        exp(rnorm(n, -0.02, 0.10))
    ),
    atinc = totinc - inctax + transfers
  )
```

Five people are not enough to do economics with. For the rest of this
page we will use a larger dataset called `cis`, containing 500
working-age people with their age, sex, education, total income,
government transfers received, and income tax paid. It is artificial,
but it is built to behave like the Canadian Income Survey extracts you
will use in the assignments, and it is already loaded here.

### Reading data from a file

In an assignment you would not be handed a dataset — you would read one
in from a file you downloaded. The `cis` data on this page is the file
`ec313cis.csv` that you downloaded at the start. Put the file in the
same folder as your Quarto document, then read it with the function
matching its format:

```{r}
#| eval: false
# comma-separated values
cis <- read_csv("ec313cis.csv")

# Stata format (a .dta file), using the haven package
library(haven)
cis <- read_dta("filename.dta")
```

Reading a file gives you a data frame, which you then work with exactly
as we are about to.

::: callout-warning
## "Cannot open file"

This error nearly always means R is looking in a different folder than
you are. R looks in its **working directory**. The reliable fix is to
create an RStudio **Project** for the course (File → New Project) and
keep your data and documents inside that project folder. Then paths
just work.
:::

### Looking at your data

Never start analysing data you have not looked at. A handful of
functions do most of the work.

`glimpse()` gives you the shape of the dataset and the type of every
column:

```{r}
glimpse(cis)
```

`head()` shows the first few rows, so you can see what actual values
look like:

```{r}
head(cis)
```

You will notice the word **tibble** in that output. A tibble is the
tidyverse's version of a data frame: the same rectangle of columns, but
printed more helpfully — it shows you the type of each column and stops
after ten rows instead of filling your screen. Anything that works on a
data frame works on a tibble, so you can treat the two words as
interchangeable.

And `summary()` describes a single variable — remember that
`cis$totinc` is just a vector:

```{r}
summary(cis$totinc)
```

For a categorical variable, count the categories instead:

```{r}
count(cis, educ)
```

Which summary you use depends on the kind of variable, and the
distinction matters:

-   A **continuous** variable can take any value in a range — income,
    tax paid, age. Averages and percentiles make sense, so use
    `summary()`.
-   A **categorical** variable sorts people into groups — education,
    province, sex. The groups are often stored as numbers, but those
    numbers are labels, not quantities. The average of "1 = man, 2 =
    woman" means nothing. Count the categories with `count()` instead.

#### The data viewer

The Console is not the only way to look. RStudio has a spreadsheet-style
viewer built in. Click the name `cis` in the **Environment** pane, or
run:

```{r}
#| eval: false
View(cis)
```

Note the capital `V` — `view()` will not work. Either way, the dataset
opens in a tab beside your document, where you can scroll through all
500 rows, sort by clicking a column heading, and filter to particular
values.

This is often the fastest way to get a feel for a dataset you have never
seen before, and it is worth doing alongside `glimpse()` rather than
instead of it. Two things to know: nothing you do in the viewer changes
your data or gets recorded in your document, so you cannot break
anything by exploring; and for the same reason, sorting or filtering
there is no substitute for writing the code. Only the code goes into
your assignment.

::: callout-tip
Look at a dataset this way every time you open one, before doing
anything else. Most analysis mistakes are really misunderstandings
about what is in the data.
:::

### Missing values in real data

Survey files rarely mark missing values with a tidy `NA`. It is common
to use an impossible number instead — in the Canadian Income Survey,
the survey your assignment extracts come from, missing earnings are
stored as 999999999996. R has no way to know that this is not a real
income:

```{r}
earnings_raw <- c(45000, 62000, 999999999996, 38000)
mean(earnings_raw)
```

An average income of 250 billion dollars is a clue that something is
wrong, but a subtler code would slip through unnoticed. Convert the code
to `NA` first, with `na_if()`:

```{r}
earnings_clean <- na_if(earnings_raw, 999999999996)
earnings_clean
mean(earnings_clean, na.rm = TRUE)
```

Now the mean describes the three people we actually have data for.

::: callout-warning
## Read the documentation first

Every survey comes with documentation — a codebook listing what each
variable measures and how missing values are coded. Read it before you
compute anything. A missing-value code left in the data can invalidate
every number in your assignment, and nothing in R will warn you.
:::

## Manipulating data

Five verbs cover most of what you will do. Each takes a data frame and
returns a modified data frame.

### The pipe

Before the verbs themselves, one piece of punctuation. You are about to
see the symbol `|>` everywhere. It is the **pipe**, and it takes the
thing on its left and feeds it into the function on its right. These two
lines do exactly the same thing:

```{r}
mean(incomes)
incomes |> mean()
```

For a single function the pipe is not worth the trouble. Its value is in
what comes next: when several operations happen one after another, the
pipe lets you read them left to right, in the order they actually
happen, instead of from the inside out.

### `filter()` — keep some rows

How many people in the sample paid no income tax at all?

```{r}
cis |>
  filter(inctax == 0) |>
  nrow()
```

Note the double equals `==`. A single `=` assigns; a double `==` tests
for equality. Conditions can be combined with `&` (and) and `|` (or):

```{r}
cis |>
  filter(educ == "University" & age < 30) |>
  nrow()
```

`|` keeps a row if either condition holds. To count everyone with a
college or university education:

```{r}
cis |>
  filter(educ == "College" | educ == "University") |>
  nrow()
```

Writing `&` here would be a mistake: nobody's education is both college
*and* university, so it would keep no one.

### `select()` — keep some columns

```{r}
cis |>
  select(id, educ, totinc, inctax) |>
  head(3)
```

### `mutate()` — create variables

The **average tax rate** — tax paid as a share of income — is the
workhorse variable of tax analysis, and it is not in the data. Create
it:

```{r}
cis <- cis |>
  mutate(atr = inctax / totinc)
```

Notice that the result is saved back into `cis` with `cis <- cis |>`.
Without that, R shows you the new variable but does not keep it.

`if_else()` creates a variable that takes one value where a condition is
true and another where it is false. It takes three arguments: the
condition, the value where it is `TRUE`, and the value where it is
`FALSE`. Either value can be a calculation. One `mutate()` can create
several variables, and a variable created on one line can be used on the
lines below it. Here is an imaginary tax credit worth 5% of income for
people under 30 with incomes below \$40,000:

```{r}
cis <- cis |>
  mutate(
    young  = as.numeric(age < 30),
    credit = if_else(young == 1 & totinc < 40000, 0.05 * totinc, 0)
  )

cis |>
  select(age, totinc, young, credit) |>
  head(5)
```

### `summarise()` — boil the data down

`summarise()` collapses a whole dataset into a single row of summary
numbers. You name each thing you want, and say how to calculate it:

```{r}
cis |>
  summarise(
    n        = n(),
    mean_inc = mean(totinc),
    mean_atr = mean(atr)
  )
```

Three numbers describing all 500 people: how many there are, their
average income, and the average share of income they pay in tax. `n()`
is a special helper that counts rows.

Note what happened to the shape of things. `filter()`, `select()` and
`mutate()` all handed back a dataset with one row per person.
`summarise()` hands back one row, full stop — the people are gone, and
what remains is a description of them.

### `group_by()` — do it separately for each group

A single average for everyone is rarely what you want. The interesting
question is almost always comparative: do incomes and tax rates differ
*by education*, *by sex*, *by province*?

`group_by()` answers that. On its own it changes nothing you can see —
it simply marks the dataset as divided into groups. The work happens
when `summarise()` runs afterwards, because it now produces one row per
group rather than one row overall:

```{r}
cis |>
  group_by(educ) |>
  summarise(
    n        = n(),
    mean_inc = mean(totinc),
    mean_atr = mean(atr)
  )
```

That is the identical `summarise()` call as before, with one line added
above it — and the result is already a public finance finding: income
rises with education, and the average tax rate rises with it. People
with higher incomes pay a larger *share* of their income in tax, which
is what it means for an income tax to be **progressive**.

Group by something else and you ask a different question. Here is the
raw gender income gap:

```{r}
cis |>
  group_by(female) |>
  summarise(mean_inc = mean(totinc), n = n())
```

This pairing is the workhorse of applied work. Most tables you will ever
produce are some version of "group by one thing, summarise another."

Square brackets work inside `summarise()` too. There they pick out the
values in the rows that meet a condition, within each group. This
calculates the gender income gap separately for each level of
education, along with the highest income in each:

```{r}
cis |>
  group_by(educ) |>
  summarise(
    gap     = mean(totinc[female == 0]) - mean(totinc[female == 1]),
    max_inc = max(totinc)
  )
```

`gap` is the average income of men minus the average income of women,
and `max()` gives the largest value.

::: callout-note
## Ordering categories

Notice that education came out alphabetically, which is not the natural
order. To fix that, make it a **factor** with the levels you want:

```{r}
cis <- cis |>
  mutate(educ = factor(
    educ,
    levels = c("Less than HS", "High school", "College", "University")
  ))

count(cis, educ)
```
:::

## Graphs

Graphs in R are built with **ggplot2**, loaded as part of the tidyverse.
Every plot has the same three ingredients: the data, a mapping from
variables to visual properties (`aes`), and a geometry (`geom_`) saying
what to draw.

```{r}
#| label: fig-inc-hist
#| fig-cap: "Distribution of total income"
#| fig-width: 6
#| fig-height: 3.5
ggplot(cis, aes(x = totinc)) +
  geom_histogram(bins = 30) +
  labs(x = "Total income ($)", y = "Number of people")
```

The long right tail — most people bunched at moderate incomes, a few
far out to the right — is characteristic of income data, and you will
see it in every income survey you ever open.

Note that ggplot layers are joined with `+`, not the pipe. This is an
historical quirk and a common source of confusion.

Mapping a variable to colour or to the x axis of a boxplot lets you
compare groups:

```{r}
#| label: fig-atr-educ
#| fig-cap: "Average tax rates by education level"
#| fig-width: 6
#| fig-height: 3.5
ggplot(cis, aes(x = educ, y = atr)) +
  geom_boxplot() +
  labs(x = NULL, y = "Average tax rate")
```

And a scatterplot shows how two continuous variables move together.
Adding `geom_smooth(method = "lm")` draws the regression line through
the points:

```{r}
#| label: fig-inc-tax
#| fig-cap: "Income tax paid against total income"
#| fig-width: 6
#| fig-height: 3.5
#| message: false
ggplot(cis, aes(x = totinc, y = inctax)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(x = "Total income ($)", y = "Income tax paid ($)")
```

The upward slope is the tax schedule at work: more income, more tax.
With this many points the cloud is dense, which is normal for survey
microdata. The line is what carries the message.

### Line graphs and groups

A line graph joins points from left to right, which suits something
that changes along the horizontal axis, such as time. To draw a separate
line for each group, map the group to `colour` inside `aes()`. First,
calculate average income by years of schooling, separately for men and
women:

```{r}
#| message: false
inc_school <- cis |>
  group_by(yrschool, female) |>
  summarise(mean_inc = mean(totinc))
```

Then draw one line for each:

```{r}
#| label: fig-inc-school
#| fig-cap: "Average income by years of schooling"
#| fig-width: 6
#| fig-height: 3.5
ggplot(inc_school, aes(x = yrschool, y = mean_inc, colour = factor(female))) +
  geom_line() +
  geom_vline(xintercept = 12, linetype = "dashed") +
  scale_colour_discrete(labels = c("Men", "Women")) +
  labs(x = "Years of schooling", y = "Average income ($)", colour = NULL)
```

Three things to notice:

-   `colour = factor(female)` draws one line for each group, in its own
    colour. `factor()` matters: `female` is a number, 0 or 1, and
    without `factor()` R treats it as a quantity rather than as two
    groups, and draws a single line.
-   `geom_vline()` draws a vertical line where `xintercept` says, here
    at 12 years of schooling, the end of high school. When the
    horizontal axis holds dates, give `xintercept` a date, such as
    `as.Date("2025-04-30")`.
-   `scale_colour_discrete(labels = ...)` replaces the 0 and 1 in the
    legend with words, in the same order. In `labs()`, `colour = NULL`
    removes the legend's title.

::: callout-important
## Assignment graphs need labels

Default axis labels are variable names, which are not acceptable in
submitted work. Always set `labs()` with readable labels and units.
:::

## Regression

Linear regression uses `lm()`, with a formula written
`outcome ~ predictor1 + predictor2`:

```{r}
model1 <- lm(inctax ~ totinc, data = cis)
summary(model1)
```

That is a lot of output, but only a few parts matter:

-   **`Estimate`** — the coefficient. Here it says that an extra dollar
    of income raises tax paid by about 34 cents on average — an
    estimate of the effective marginal tax rate in this sample.
-   **`Std. Error`** — how precisely the coefficient is estimated.
-   **`Pr(>|t|)`** — the p-value for the test that the coefficient is
    zero. Small means the estimate is distinguishable from no effect.
-   **`Multiple R-squared`** — the share of the variation in tax paid
    the model explains.

A second regression asks a different question: is the tax system
progressive? If the average tax rate rises with income, it is —
that is essentially the definition:

```{r}
cis <- cis |>
  mutate(lntotinc = log(totinc))

model2 <- lm(atr ~ lntotinc, data = cis)
```

The positive coefficient on log income confirms that people with higher
incomes pay a larger share of their income in tax.

### Categorical predictors

You can put a categorical variable straight into a regression, and R
creates the dummy variables for you:

```{r}
model3 <- lm(atr ~ educ, data = cis)
coef(model3)
```

Notice there are only three coefficients for four education categories.
R leaves one out — here "Less than HS" — as the **base group**, and
every other coefficient says how that group compares to it. There is no
way to include all four; the base group is what the others are measured
against.

`coef()` was used above to show just the coefficients. It returns a
named vector, which is the same kind of object from the start of this
page. To keep just one coefficient, put its name in square brackets.
This stores the marginal tax rate from the first regression as a single
number you can calculate with:

```{r}
mtr <- coef(model1)["totinc"]
mtr
```

To compare against a different group, change which level comes first:

```{r}
cis_hs <- mutate(cis, educ = relevel(educ, ref = "High school"))
model4 <- lm(atr ~ educ, data = cis_hs)
coef(model4)
```

Now everything is relative to high school graduates, and the "Less than
HS" coefficient is negative, as you would expect from a progressive
tax paired with lower incomes.

### Presenting results

Console output is not acceptable in submitted work. Use
**modelsummary** to produce a proper table:

```{r}
#| message: false
library(modelsummary)

modelsummary(
  list("Tax schedule" = model1, "Progressivity" = model2),
  stars = TRUE
)
```

This is the format your assignments should use.

Two arguments help when a regression has many dummy variables.
`coef_omit` leaves out every coefficient whose name contains the text
you give it, and `gof_map` chooses which statistics appear at the
bottom. Here the education dummies are left out:

```{r}
model5 <- lm(atr ~ lntotinc + educ, data = cis)

modelsummary(
  list("Without education" = model2, "With education" = model5),
  coef_omit = "educ", gof_map = c("nobs", "r.squared")
)
```

To leave out more than one kind of coefficient, separate them with `|`,
as in `coef_omit = "educ|female"`.

### Tables of data

`modelsummary()` is for regressions. For tables of data, the same
package has two more functions.

`datasummary_df()` turns a data frame, such as one you have made with
`summarise()`, into a table. `fmt` sets the number of decimal places:

```{r}
cis |>
  group_by(educ) |>
  summarise(mean_inc = mean(totinc), mean_atr = mean(atr)) |>
  datasummary_df(fmt = 2)
```

`datasummary()` builds a table from a formula, `rows ~ columns`. Here
each level of education is a row, and the column is the mean of
`totinc`:

```{r}
datasummary(educ ~ totinc * Mean, data = cis, fmt = 0)
```

The column is headed "Mean", which is not very informative. `Heading()`
replaces the heading of whatever comes after it:

```{r}
datasummary(educ ~ totinc * Heading("Average income ($)") * Mean,
            data = cis, fmt = 0)
```

Give `datasummary()` its data with `data =`, as above. Its first
argument is the formula, so piping a data frame into it does not work.

# Part 2: Quarto

Part 1 was about getting answers. This part is about presenting them.
Quarto is how you turn the code and results from Part 1 into the PDF you
hand in.

## What a Quarto document is

A Quarto file (extension `.qmd`) is a single document containing your
writing, your code, and the output of that code. When you **render**
it, Quarto runs every piece of code fresh and produces a finished PDF
with the results embedded.

This matters for two reasons. Your work is reproducible, since the
document cannot drift out of step with the code that produced it. And
you never have to copy and paste a number or a graph again — a constant
source of errors.

To create one in RStudio: File → New File → Quarto Document.

::: callout-note
## Download the practice file

Rather than starting from a blank document, start from this one. It is
a working Quarto file containing everything on this page: it loads
packages, reads in the practice data, summarises it, draws a graph, and
runs a regression. If you downloaded both files at the start, you
already have them.

[{{< fa solid file-export >}} `ec313-r-template.qmd`](/files/tutorials/ec313-r-template.qmd){download="ec313-r-template.qmd"}

[{{< fa solid file-export >}} `ec313cis.csv`](/files/tutorials/ec313cis.csv){download="ec313cis.csv"}

Download **both** files into the same folder, open the `.qmd` in
RStudio, and press **Render**. You should get a PDF. Then start changing
things — that is how you learn this.

It uses the same `cis` data as this page, read in with `read_csv()`, so
every result you see above you can reproduce and alter yourself.
:::

## The three ingredients

A `.qmd` file has exactly three kinds of content.

**1. The YAML header**, at the very top, between three dashes. It sets
the title and output format:

``` yaml
---
title: "EC313 Assignment 1"
author: "Your Name, 123456789"
format: pdf
---
```

**2. Text**, written in Markdown. A few marks do most of the work:

| You type | You get |
|:------------------------|:----------------------------------------|
| `## Question 1` | A section heading |
| `**important**` | **important** |
| `*emphasis*` | *emphasis* |
| `- item` | A bullet point |
| `$\beta_1$` | An inline equation, $\beta_1$ |

**3. Code chunks**, fenced by three backticks with `{r}`. In RStudio the
keyboard shortcut is Ctrl+Alt+I (Cmd+Option+I on a Mac):

````{verbatim}
```{r}
mean(cis$totinc)
```
````

Chunk options go on lines starting with `#|` inside the chunk. The ones
you will want:

| Option | Effect |
|:------------------|:----------------------------------------------------|
| `#| echo: false` | Run the code but hide it; show only output |
| `#| eval: false` | Show the code but do not run it |
| `#| message: false` | Suppress package startup chatter |
| `#| warning: false` | Suppress warnings |
| `#| fig-cap: "..."` | Add a caption to a figure |

For this course, **leave code visible** — showing your code is part of
what is being marked.

## Inline code

You can also put a computed number directly into a sentence, rather than
typing it by hand. In your text, write this:

````{verbatim}
The mean total income is `r round(mean(cis$totinc))` dollars.
````

and when you render, R replaces it with the answer:

> The mean total income is `r round(mean(cis$totinc))` dollars.

That number was computed while this page was being built. If the data
changed, the sentence would change with it — which is why you should
never copy a number out of your output and retype it.

## Rendering

Everything you hand in for this course is a PDF, and producing a PDF
needs a LaTeX installation working quietly in the background. You almost
certainly do not have one yet. Install it once, in the Console, and then
never think about it again:

```{r}
#| eval: false
install.packages("tinytex")
tinytex::install_tinytex()
```

It takes a few minutes and downloads a few hundred megabytes. If you
prefer, the same thing can be done from the Terminal tab with
`quarto install tinytex`.

::: callout-warning
Do this now, not later. The first time most people try to produce a PDF
is an hour before a deadline, and that is a bad moment to discover a
missing LaTeX installation. If a render fails with an error mentioning
LaTeX, `tlmgr`, or a missing `.sty` file, this is what it is telling
you.
:::

With that in place, press the **Render** button at the top of the Source
pane, or Ctrl+Shift+K (Cmd+Shift+K on a Mac). Quarto runs your code from
top to bottom in a clean session and writes a PDF next to your `.qmd`.

::: callout-warning
## Render early and often

Because rendering starts from a clean session, a document can fail to
render even though everything worked in your Console — usually because
an object was created in the Console and never written into the
document. Render after finishing each question. Discovering a problem at
9:00pm on the due date is avoidable.
:::

## Common errors

Everyone hits these. They are not a sign that you are bad at this.

| Message | Usual cause |
|:-----------------------------|:-----------------------------------------|
| `could not find function "x"` | A package is not loaded — add the `library()` call |
| `object 'x' not found` | Typo, wrong capitalisation, or the object was never created |
| `unexpected symbol` | A missing comma, bracket, or quotation mark |
| `cannot open file` | Wrong working directory — use an RStudio Project |
| A `+` in the Console | R is waiting for you to finish an incomplete command; press Esc |

::: callout-tip
## How to read an error

Errors look intimidating but usually name the problem in the first
line. Read that line, look at the line of code it points to, and check
the simple things first: spelling, capitalisation, and matching
brackets. Pasting the message into a search engine is a legitimate and
widely used technique.
:::

## Where to get help

Everything below is free to read online. You do not need to buy anything
for this course beyond the textbook.

-   [R for Data Science](https://r4ds.hadley.nz) — the standard
    reference for the tidyverse, free in full on that site. Chapters 1
    through 5 cover most of what this course needs.
-   [Posit Cheatsheets](https://posit.co/resources/cheatsheets/) —
    free one-page references for dplyr, ggplot2, and RStudio itself.
    Print the dplyr and ggplot2 ones.
-   [The Quarto guide](https://quarto.org/docs/guide/) — official
    documentation, well organised and full of examples.
-   [ggplot2 reference](https://ggplot2.tidyverse.org/reference/) — every
    geometry, with pictures.
-   Office hours. If you are stuck for more than twenty minutes on
    something mechanical, bring it to me rather than losing an evening
    to it.
