---
title: "EC313 Assignment 1: Who Paid the Carbon Tax at the Pump?"
author: "Your name, your student number"
date: today
format:
  pdf:
    fig-width: 6
    fig-height: 3.5
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{table}{H}
        \floatplacement{figure}{H}
execute:
  message: false
  warning: false
---

<!--
  HOW TO USE THIS FILE
  --------------------
  1. Put this file and ec313assign1.csv in the SAME folder.
  2. Open this file in RStudio and press Render. It should produce a PDF
     before you have written anything. If it does not, fix that first.
  3. Write your code in the chunks provided and your answers where it
     says "Write your answer here." Render after every question.
  4. When you are finished, upload THIS .qmd file to the Gradescope
     assignment "Assignment 1 - Code", and the PDF it produces to
     "Assignment 1 - Written".

  Some answers are checked by computer. The computer looks for the
  objects named in each question, so use exactly those names, and do not
  change the code in the "setup" and "load-data" chunks.
-->

## Background

From 2019 to 2025, the federal government charged a carbon price on
fuel in provinces that did not have a carbon pricing system of their
own that met the federal standard. The charge was paid to the
government by fuel distributors, not by drivers, and from April 2024 it
was 17.61 cents per litre of gasoline.
British Columbia had its own carbon tax at the same rate. Quebec prices
carbon through a cap-and-trade system instead, with no separate charge
per litre at the pump. Most of the revenue from the federal charge was
returned to households as the Canada Carbon Rebate, a payment that did
not depend on how much fuel a household bought.

On April 1, 2025, the federal charge was set to zero, and British
Columbia ended its carbon tax on the same day. Nothing changed in
Quebec.

Who bore the carbon tax at the pump? The theory of tax incidence says
that the answer depends on the elasticities of supply and demand, not on
who sends the money to the government. If sellers had passed the whole
tax on to drivers, the price of gasoline should have fallen by the full
amount of the tax when it was removed. If sellers had absorbed part of
the tax themselves, the price should have fallen by less. In this
assignment you will measure how much the price fell, and use that
measurement to work out who bore the tax.

## The data

The data come from Statistics Canada's monthly survey of retail
gasoline prices (Table 18-10-0001-01). They are the average price of
regular unleaded gasoline at self-service stations in 15 cities, for
each month from October 2024 to September 2025: six months before the
carbon tax was removed, and six months after.

| Variable   | Description                                                                  |
|:-----------|:-----------------------------------------------------------------------------|
| `city`     | City                                                                         |
| `province` | Province                                                                     |
| `date`     | The month, stored as a date (the first day of the month)                     |
| `month`    | The month, as text, such as "2025-03" for March 2025                         |
| `gas`      | Average price of regular gasoline, in cents per litre, including all taxes   |
| `salestax` | The GST or HST rate charged on gasoline in the province (plus the QST in Quebec), as a proportion: 0.13 means 13% |

The GST and HST are charged on the price of gasoline *including* the
carbon charge, so removing the 17.61 cent charge removes
17.61 × (1 + `salestax`) cents from the price at the pump.

Winnipeg is left out, because Manitoba's own gas tax went back up by
12.5 cents in January 2025, in the middle of the period. Whitehorse and
Yellowknife are left out too, because prices there come from few
stations and often do not change for months at a time. In Halifax, the
HST also fell from 15% to 14% on April 1, 2025, which lowered the price
there by about one cent more.

Source: Statistics Canada, Table 18-10-0001-01, Monthly average retail
prices for gasoline and fuel oil, by geography, October 2024 to
September 2025. Reproduced and distributed on an "as is" basis with the
permission of Statistics Canada.

## Instructions

-   Write your R code in the chunks provided. You can add more chunks if
    you need them.
-   When a question tells you to store a result under a particular name,
    such as `q2` or `pt_naive`, use exactly that name. Those objects are
    marked by computer, and a different name counts as missing.
-   Do not round the numbers you store. Round only when you display them.
-   Tables and graphs must be presentable: use `modelsummary()`,
    `datasummary()` or `datasummary_df()` for tables, and give every
    graph readable axis labels.
-   The hints under each question point to sections of the R tutorial,
    the Introduction to R and Quarto page on the course website. Each
    section has a worked example of the step the hint describes, using
    different data.
-   Render the document after each question, not only at the end.
-   Every written answer goes under **Answer:**, and what it must cover is
    set out just above, under **Comment on the results**.
-   Each question is marked out of 5 from your PDF, using the rubric on
    the assignment page. The objects you are told to create are also
    checked by computer, for another 5 marks, so the assignment is out
    of 30.

## Setup

Do not change this chunk or the next one.

```{r}
#| label: setup
library(tidyverse)
library(modelsummary)
```

```{r}
#| label: load-data
gas <- read_csv("ec313assign1.csv")

glimpse(gas)
```

## Question 1 (5 marks)

Create the following variables and add them to `gas`:

-   `removed`: 1 if the city's carbon tax was removed on April 1, 2025
    (every city outside Quebec), 0 for the two Quebec cities
-   `post`: 1 for April 2025 and every month after it, 0 for the months
    before
-   `carbon`: the carbon tax included in the price, in cents per litre,
    counting the sales tax charged on it. It is 17.61 × (1 + `salestax`)
    in cities where the tax was removed, in the months before April 2025.
    It is 0 in every other case: in those cities from April 2025 on, and
    in the Quebec cities in every month.

Then make a table showing the value of `carbon` in each of the 15
cities in March 2025, the last month before the tax was removed.

*Hints.*

-   Create all three variables with `mutate()`, and save the result back
    into `gas`; otherwise the new variables are not kept. Inside one
    `mutate()`, a variable you have just created can be used in the
    lines below it. See "`mutate()` — create variables" in the R
    tutorial.
-   For `removed`, a comparison such as `province == "Quebec"` gives
    `TRUE` or `FALSE` for every row, and `!=` means "is not equal to".
    `as.numeric()` turns `TRUE` and `FALSE` into 1 and 0. See "Picking
    out elements" and "Vectors have a type" in the R tutorial.
-   For `post`, comparisons such as `>=` also work on dates. To compare
    `date` with April 1, 2025, write that date as
    `as.Date("2025-04-01")`. See "Dates" in the R tutorial.
-   For `carbon`, `if_else(condition, a, b)` gives `a` in the rows where
    the condition is `TRUE` and `b` in the rows where it is `FALSE`. A
    condition can combine two others with `&` (and), such as one about
    `removed` and one about `post`. In R, multiplication is `*`, not ×.
    See "`mutate()` — create variables" in the R tutorial.
-   For the table, use `datasummary(city ~ carbon * Mean, data = ...)`,
    with `data` set to the March 2025 rows of `gas` (use `filter()`).
    Each city has only one row in March, so its mean is just its March
    value. Give the column a clearer heading than "Mean". See "Tables of
    data" in the R tutorial.

```{r}
#| label: q1-variables
# Add removed, post and carbon to gas here.

```

```{r}
#| label: q1-table

```

**Comment on the results.** Use the theory of tax incidence from class
to predict what should have happened to the price of gasoline in
Toronto when the tax was removed. By how much should the price have
fallen if the supply of gasoline were perfectly elastic? By how much if
it were perfectly inelastic? Which of the two do you expect to be closer
to what happened, and why?

**Answer:**

*Write your answer here.*

## Question 2 (5 marks)

For each month, calculate the average price of gasoline in the cities
where the carbon tax was removed, and the average in the two Quebec
cities. That gives 24 averages: 12 months, for each of the two groups.
Store the result in a data frame called `q2`, with the averages in a
column called `avg_gas`.

Then make a line graph of the two averages over time, with the date on
the horizontal axis, one line for each group, and a vertical line
marking April 2025.

*Hints.*

-   For the averages, use `group_by(date, removed)` and then
    `summarise()`. Grouping by both variables gives a separate average
    for each month in each group: the 24 rows. See "`group_by()` — do it
    separately for each group" in the R tutorial.
-   You will need the averages for your comment. To see them, run the
    chunk (the green arrow at its top right), then type `q2` in the
    Console.
-   For the graph, start with `ggplot(q2, aes(x = date, y = avg_gas))`
    and add `geom_line()`. See "Line graphs and groups" in the R
    tutorial, which has a graph like this one.
-   Add `colour = factor(removed)` inside `aes()`. This gives each group
    its own line. Without `factor()`, R treats `removed` as a number
    rather than as two groups, and draws a single zigzag line.
-   Add `geom_vline(xintercept = as.Date("2025-04-01"), linetype = "dashed")`
    to put in the vertical line at April 2025.
-   Use `scale_colour_discrete()` to replace the 0 and 1 in the legend
    with words, giving the labels in the order 0, then 1, and use
    `labs()` to give the axes readable titles. See "Line graphs and
    groups" in the R tutorial.

```{r}
#| label: q2-averages

```

```{r}
#| label: q2-graph

```

**Comment on the results.** Describe how the two lines move before
April 2025, and what happens to the gap between them in April 2025. How
does the change in the gap compare with the size of the carbon tax you
found in Question 1?

**Answer:**

*Write your answer here.*

## Question 3 (5 marks)

The simplest way to measure how much of the tax cut reached drivers is
to compare the price just before the tax was removed with the price
just after.

For each city, calculate the change in the price of gasoline from March
2025 to April 2025 (the April price minus the March price, so a fall is
negative). Store the result in a data frame called `q3`, with one row
per city and columns called `city`, `removed`, `carbon` (the tax in
March) and `change`. Present `q3` as a table.

Then calculate the **pass-through rate** for each of the 13 cities where
the tax was removed: the fall in price divided by the tax removed, or
`-change / carbon`. A pass-through rate of 1 means the price fell by
exactly the amount of the tax. Store the average of the 13 rates, as a
single number, in `pt_naive`.

*Hints.*

-   Keep only March and April 2025 with `filter()`. A row cannot be in
    both months at once, so combine the two conditions with `|` (or),
    not `&`. See "`filter()` — keep some rows" in the R tutorial.
-   Then group by `city` and `removed` (grouping by `removed` as well
    keeps it as a column in `q3`), and use `summarise()` to create
    `carbon` and `change`.
-   Inside `summarise()`, square brackets pick out the values in the
    rows that meet a condition: `gas[month == "2025-04"]` is the city's
    April price. See the income gap example under "`group_by()` — do it
    separately for each group" in the R tutorial.
-   `carbon` is positive in March and 0 in April, so `max(carbon)` gives
    the tax in March.
-   `datasummary_df()` turns a data frame into a table, and `fmt = 2`
    shows the numbers in it to two decimal places. See "Tables of data"
    in the R tutorial.
-   For `pt_naive`, save the rows of `q3` where `removed == 1` as a new
    data frame. Calculate the 13 rates from its columns, which you can
    pick out with `$`, and take their `mean()`. See "Columns are
    vectors" in the R tutorial.

```{r}
#| label: q3-table

```

```{r}
#| label: q3-passthrough

```

**Comment on the results.** Report `pt_naive` and explain what it
means. What happened to the price in Montreal and Quebec City over the
same month, when their taxes did not change? What does that tell you
about using the before-and-after change to measure pass-through?

**Answer:**

*Write your answer here.*

## Question 4 (5 marks)

The price of gasoline changes from month to month for many reasons
other than taxes, such as the price of crude oil, the exchange rate and
the season. A regression with **fixed effects** separates the effect of
the tax from those other changes:

$$
gas_{it} = \beta_0 + \beta_1 carbon_{it} + \text{city dummies} + \text{month dummies} + \epsilon_{it}
$$

where $i$ is a city and $t$ is a month. The city dummies absorb
differences between cities that stay the same over time, such as
provincial fuel taxes and transport costs. The month dummies absorb
changes that hit every city at the same time, such as a fall in the
price of crude oil. What is left to estimate $\beta_1$ is how prices
changed in the cities whose tax changed *compared with* the Quebec
cities, whose tax did not. $\beta_1$ is the pass-through rate.

Because `city` and `month` are text, R creates the dummies for you when
you put them in the formula.

**(a)** Keep only March and April 2025, and estimate
`lm(gas ~ carbon + city + month, data = ...)` on those two months.
Store the regression as `q4_model`, and store the coefficient on
`carbon`, as a single number, in `pt_fe`.

*Hints.*

-   Save the March and April rows as a new data frame with `filter()`,
    as in Question 3, and use it as the `data`.
-   `coef()` gives all the coefficients of a regression. To keep just
    one, put its name in square brackets after it. See "Categorical
    predictors" in the R tutorial.

```{r}
#| label: q4a

```

**(b)** Estimate the same regression using all 12 months, and store it
as `q4_all`. Present both regressions in one table, one in each column,
without the city and month dummies.

*Hints.*

-   `modelsummary()` takes a list of regressions, and the names you give
    them become the column headings. See "Presenting results" in the R
    tutorial.
-   `coef_omit = "city|month"` leaves the city and month dummies out of
    the table, which would otherwise run to 25 rows, and
    `gof_map = c("nobs", "r.squared")` keeps just the number of
    observations and the R² at the bottom.

```{r}
#| label: q4b

```

**Comment on the results.** Interpret the coefficient on `carbon` in
the March-April regression, and compare it with `pt_naive` from
Question 3. Why are they different? Then explain why the estimate from
all 12 months is different again. Look at your graph from Question 2:
does the gap between the two lines stay the same from October 2024 to
March 2025? Which of the two estimates do you find more convincing, and
why?

**Answer:**

*Write your answer here.*

## Question 5 (5 marks)

In class we found that a unit tax $t$ raises the price consumers pay by

$$
\Delta P_c = \frac{\varepsilon_s}{\varepsilon_s + |\varepsilon_d|}\, t
$$

so the pass-through rate is
$\theta = \varepsilon_s / (\varepsilon_s + |\varepsilon_d|)$.

**(a)** Rearrange this equation to express the elasticity of supply,
$\varepsilon_s$, in terms of $\theta$ and $|\varepsilon_d|$, and show
your steps under **Answer:** below. Then take the short-run elasticity of demand
for gasoline to be $-0.2$, and calculate the elasticity of supply
implied by `pt_fe`, your estimate from Question 4(a). Store it, as a
single number, in `eps_s`.

*Hint.* Calculate `eps_s` from `pt_fe` itself, not from a number copied
out of your table. The table rounds `pt_fe`, and because `pt_fe` is
close to 1, even a small rounding error changes `eps_s` a lot.

```{r}
#| label: q5a

```

**(b)** **Comment on the results.** Who bore the economic burden of the
carbon tax at the pump? In your answer, consider:

-   the value of `eps_s`, and why the supply of gasoline to a single
    province might be this elastic. How would `eps_s` change if you used
    the all-months estimate from Question 4(b) instead?
-   the charge was paid to the government by fuel distributors, not by
    drivers. Does that change who bore it?
-   the Canada Carbon Rebate ended along with the tax. A household's
    rebate depended on its province, its size and whether it lived in a
    rural area, but not on how much fuel it bought, while the tax it
    paid rose with every litre. Given your answer about who bore the
    tax, which households were better off when the tax and the rebate
    ended together, and which were worse off?

**Answer:**

*Write your answer here.*

## Use of generative AI

If you used a generative AI tool for any part of this assignment, name
the tool and describe what you used it for. If you did not use one, say
so.

*Write your statement here.*

## Before you submit

This chunk lists any required objects that are still missing. It should
say "All required objects found" before you submit.

```{r}
#| label: check
#| echo: false
required <- c("gas", "q2", "q3", "pt_naive", "q4_model", "pt_fe", "q4_all", "eps_s")
missing  <- required[!sapply(required, exists)]
if (exists("gas")) {
  missing_cols <- setdiff(c("removed", "post", "carbon"), names(gas))
  if (length(missing_cols) > 0) missing <- c(missing, paste0("gas$", missing_cols))
}
if (length(missing) == 0) {
  cat("All required objects found\n")
} else {
  cat("Still missing:", missing, sep = "\n  ")
}
```
